---
title: "Passeriformes diversification"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This article aims to...

First of all, to run the analysis within this vignette we'll need to load the `treesliceR`, `ape` and `ggplot2` packages.
```{r setup}

libs <- c("ape", "ggplot2")
if (!requireNamespace(libs, quietly = TRUE)){
    install.packages(libs)
  }
library(ape)
library(ggplot2)

libs <- c("treesliceR")
if (!requireNamespace(libs, quietly = TRUE)){
    devtools::install_github("AraujoMat/treesliceR")
  }
#devtools::install_github("AraujoMat/treesliceR") # If it isn't installed within your PC
library(treesliceR)
```

# Case study using a single passeriformes tree

## First framework example

Herein, we'll compare the passeriformes Diversification Rates (DR) before and after Australia aridification (~33 Mya). Primarily, we need to load the passeriformes phylogeny that are available within the package. The passeriformes phylogeny was obtained from [Jetz et al. (2012)](https://www.nature.com/articles/nature11631).
```{r}
tree <- pass_trees[[1]]
```

Now, we need to split the phylogeny into two distinct portions, one representing the last 33 millions of years ago and another the cladogenesis events before this period. To capture the root slice before 33 My, we can slice our phylogeny tipwardly (i.e., from tips to root) using the `squeeze_tips()` function. Alternatively, the tip slice of the last 33 My can be obtained through the `squeeze_root()` function, which slice a phylogeny rootwardly (i.e, from root to tips).

Since tip diversification rates (DR) calculation are dependent on the node path to a given tip, we will used the argument `dropNodes = TRUE` to remove those void nodes that could bias our estimation.
:
```{r}
recent <- squeeze_root(tree = tree, time = 33, dropNodes = T)
old <- squeeze_tips(tree = tree, time = 33, dropNodes = T)
```

We can visualize our outputted to see if our algorithm worked
```{r}
par(mfrow = c(1, 3))
plot(tree, main = "Complete tree", show.tip.label = F); axisPhylo()
plot(recent, main = "Recent tree", show.tip.label = F); axisPhylo()
plot(old, main = "Old tree", show.tip.label = F); axisPhylo()
par(mfrow = c(1, 1)) # Returning to a simple 1x1 viewer
```

Then, we need to calculate the tip diversification rates (DR) separately for each phylogeny slice, while finding their difference. To calculate the DR, we can use the function `DR()` available within the `treesliceR` package:
```{r}
DR_diff <- DR(tree = recent)[,2] - DR(tree = old)[,2]
```

Now we'll pair the differences in DR by assigning their difference to a data.frame that contains our passeriformes information inside our tree object:
```{r}
tree$Species_info$DR_diff <- DR_diff
```

To calculate the mean and standard deviation difference per passeriformes family, we use the `tapply()` function to aggregate the tip-DR's by family:
```{r}
fam_DR <- tapply(tree$Species_info$DR_diff, tree$Species_info$Family, mean)
fam_DR_sd <- tapply(tree$Species_info$DR_diff, tree$Species_info$Family, sd)
```

To visualize the differences in DR among families, lets create a dataframe contaning our outputs:
```{r}
fam_df <- data.frame(Family = names(fam_DR), DR_diff = fam_DR, DR_sd = fam_DR_sd)
# Sorting based on DR value to plot
fam_df <- fam_df[order(fam_df$DR_diff),]
fam_df$Family <- factor(fam_df$Family, levels = fam_df$Family)
```

Now, we can create a graph similar to the manuscript using the package `ggplot2`:
```{r, fig.width = 7}
ggplot(fam_df, aes(x = Family, y = DR_diff,
                    ymin = DR_diff - DR_sd,
                    ymax = DR_diff + DR_sd)) +
  geom_pointrange(color = "#d90429") +
  geom_hline(yintercept = 0, linetype="dashed", color = "black") +
  coord_flip() + theme_minimal() + 
  theme(axis.title = element_text(size = 13),
        axis.text = element_text(size = 10),
        axis.line = element_line(colour = "black"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank()) +
  ylab(expression(paste(DR["recent"]-DR["past"]))) + xlab(NULL)
```


Pay attention that the final outputs could be sligly different from those obtained on the [Araujo et al. (in review)]() manuscript. However, in contrast to the study, this example was executed only for a single passeriformes phylogeny. A more complete assessment using all phylogenies could be executed in the sections below. 

## Second framework example

Herein, we'll use the Australian passeriformes to calculate the CpB-rate for both turnover and nestedness components. Firstly, we'll need to load the paired assemblages with their respective neighborhoods to run the analysis which is stored in an intern object called `pass_asb`. This object is a list of presence-absences matrices, where each lists contain a given Australian focal cell and its respective neighborhood cells. So, let's check the header of the four first columns of the first matrix within the list:
```{r}
head(pass_asb[[1]][, 1:4])
```

We can run some sensitivity analysis to find the most parsimonious number of slices to assess the CpB rate patterns. But first, we'll need to create an vector of number of slices:
```{r}
vec <- c(250, 500, 750, 1000, 1250, 1500, 1750, 2000)
```

Now we can run the sensitivity analysis and visualize its patterns. We'll do only for turnover in order to :
```{r}
sens_turn <- CpR_sensitivity(tree = tree, vec = vec, samp = 100,
                             asb = pass_asb, rate = "CpB", comp = "turnover")
```

So, we can plot our sensitivity analysis for both components using the `CpR_sensitivity_plot` funcion. We use the ggplot syntaxe to add a vertical line showing our number of slices set for our following analysis:
```{r}
# Store each graph within a respective object
CpR_sensitivity_plot(sens_turn, rate = "CpB", stc = "mean") + 
  geom_vline(xintercept = 1000, linetype="dashed", color = "black")

```


Calculating the CpB-rate from the turnover and nestedness components under a multisite approach (PS: this can take some minutes):
```{r}
# For turnover component
turn <- CpB(tree = tree, n = 1000, asb = pass_asb, comp = "turnover")
```
Finally, we can plot these CpB through time or map it. To map it, we can use our intern `AU_grid` object that represents the Australian grid map. Let's do it first for the turnover component:
```{r ggplot2, echo = FALSE}
CpR_graph(data = turn, rate = "CpB", qtl = TRUE)

```

```{r}
CpR_graph(data = turn, rate = "CpB", qtl = TRUE, map = AU_grid)
```

